---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Index insurance


## Introduction

Here we evaluate the same contract as in the previous chapter. But we use predicted yield from remote sensing instead of observed yield. That is, index insurance instead of "area_yield". 


First we predict yield for all years.

```{r, message=FALSE}
library(agrins)
mods <- readRDS("rice_models.rds")
m1 <- mods$m1
idx <- data_rice("indices")
idx$yield <- predict(m1, idx)
head(idx)
```

We combine this with the household data

```{r}
z <- readRDS("hh_rice_yield.rds")
z <- merge(z, idx[, c("zone", "year", "yield")])
head(z)
```

Now compute income with and without insurance. 

We create an insurance payout function.

```{r}
get_payout <- function(y, y_z, trigger, price){
   pmax( (trigger * y_z) - y,  0) * price
}
``` 

And compute payouts. 

```{r}
price <- 0.23
premium <- 10
trigger <- 0.9
rho <- 1.5

z$income <- z$y * price
z$payout <- get_payout(z$yield, rho, trigger, price)
z$income_with_ins <- z$income + z$payout - premium
head(z)
```

On to the certainty equivalent income. We compute it with and without insurance in one step

```{r ins06}
zz <- aggregate(z[, c("income", "income_with_ins")], z[,"zone", drop=FALSE], function(i) ce_income(i, 1.5))
```

Now compute the benefit of insurance

```{r ins066}
zz$benefit <- (zz$income_with_ins - zz$income)
zz$rel_benefit <- zz$benefit/ zz$income
head(zz)
mean(zz$benefit)
```
Does the contract pass the test?


And plot it 

```{r ins606, out.width = '100%'}
quantile(zz$benefit)
barplot(sort(zz$rel_benefit), ylab="relative benefit", xlab="zone", las=1)
```


