
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# African maize trials

Here we use data that was also used in a paper by [Lobell et al.](https://www.nature.com/nclimate/) [Nonlinear heat effects on African maize as evidenced by historical yield trials](https://dx.doi.org/10.1038/NCLIMATE1043).

## Get the data

The data is on CIMMYT's dataverse. 

https://data.cimmyt.org/dataset.xhtml?persistentId=hdl:11529/10548190

The ID is "hdl:11529/10548190" and we can download the data with the "agro" package. This package is under development. You can install it like this (if need be):


```{r}  
if (!require(agro))  source("https://install-github.me/reagro/agro")
```

Now we can use the package to download the data


```{r}  
ff <- agro::get_data_from_uri("hdl:11529/10548190", ".")
ff
```

We have two "gz" (GNU zip) files. We can g-unzip these


```{r, message=FALSE}  
print(R.utils::gunzip(ff[1], remove=FALSE, overwrite=TRUE))
print(R.utils::gunzip(ff[2], remove=FALSE, overwrite=TRUE))
```

Now we have zip files that we can unzip

```{r}  
unzip("hdl_11529_10548190/EIL_site_latlon.zip")
unzip("hdl_11529_10548190/maizedata.lobell.sep2011.csv.zip")
```

Read the data

```{r}  
location <- read.csv('EIL_site_latlon.csv', stringsAsFactors=FALSE)
maize_all <- read.csv('maizedata.lobell.sep2011.csv', stringsAsFactors=FALSE)
```

## Explore 

```{r}
head(location)
dim(maize_all)
```

```{r}
maize <- maize_all[,c(1:9)]
maize$yield <- round(exp(maize$logYield), 1)

table(maize$Management)
plot(sort(maize$yield))
```


```{r, fig.width=10}
par(mfrow=c(1,2))
boxplot(yield ~ Management, data=maize, horizontal=TRUE, las=2)
boxplot(yield ~ vargroup, data=maize, horizontal=TRUE, las=2)
```

```{r}
library(maptools)
data(wrld_simpl)
plot(location$Longitude, location$Latitude, col="red", pch=20)
plot(wrld_simpl, add=TRUE)
```

## Combine 

```{r}
d <- merge(location, maize, by.x="LocationID", by.y="sitecode")
```


```{r}
dsub <- d[, c("yield","Longitude", "Latitude", "Management", "vargroup", "Country" )]
da <- aggregate(yield ~ ., data=dsub, median)
#da <- aggregate(d[,"yield",drop=FALSE], d[,-1], data=dsub, median)
da <- da[order(da[,1], da[,2]), ]
head(da)
table(da$Management, da$vargroup)
dopt <- da[da$Management=="Optimal", ]
```

## Model

```{r, message=FALSE}
head(da)
library(randomForest)
m <- randomForest(yield ~ Longitude + Latitude, data=da)
m
p = predict(m)
plot(da$yield, p)
abline(0,1)
```

```{r}
library(raster)
e <- extent(c(-22, 60, -37, 24))
aoi <- raster(ext=e, res=1/6)
pp <- interpolate(aoi, m, xyNames=c('Longitude', 'Latitude'))
pp <- mask(pp, wrld_simpl)
plot(pp)
points(da$Longitude, da$Latitude, col="blue")

lon = init(aoi, "x")
lat = init(aoi, "y")
s <- stack(lon, lat)
names(s) <- c("Longitude", "Latitude")
p2 <- predict(s, m)
```


```{r}
bio <- getData("worldclim", var="bio", res="10")
plot(bio)
e <- extract(bio, da[, c("Longitude", "Latitude")])
de <- cbind(da[,"yield",drop=FALSE], e)
m <- randomForest(yield ~ ., data=de)
x <- predict(bio, m, ext=aoi)
plot(x)
```

```{r}
de <- cbind(da, e)
head(de)
de$Country = NULL
de$Management <- as.factor(de$Management)
de$vargroup <- as.factor(de$vargroup)
m2 <- randomForest(yield~., data=de)
m2
varImpPlot(m2)
```

```{r}
afbio <- crop(bio, s)
predictors <- stack(afbio, s)
df <- data.frame(Management="Optimal", vargroup="EPOP", stringsAsFactors = F)
df2 <- data.frame(Management="Optimal", vargroup="ILPO", stringsAsFactors = F)

#df <- data.frame(Management=5, vargroup=2)
pxd <- predict(predictors, m2, const=df2)
plot(pxd)
```

Now determine which vargroup performs best where, on average?

Do this for optimal and for drought?

when done, make it better...






