
# Explaining rice yield variability in Western Nepal

Author names

## Introduction


## Chapter requirements


## Data 


```{r}
rice_npl <- read.csv("rice_nepal_survey.csv", header=T)
rice_npl_rf <- read.csv("rice_nepal_rain.csv", header=T)

rice_rf <- merge (rice_npl, rice_npl_rf)

#First manage the soil related variables before merging with rice_rf df 
rice_soil <- read.csv("rice_nepal_soil.csv")
ls(rice_soil)

#Rice soil data need to be curated (all profiles data are same so just take the top layer data)
ls(rice_soil)
```


Conversions to standard units

```{r}
# mg/kg to %
rice_soil$SOC <-(rice_soil$ORCDRC_s1)/1000*100 
rice_soil$PH <- (rice_soil$PHIHOX_s1)/10
soil <- subset(rice_soil, select = c(hhid, SOC, PH, CLYPPT_s1, PHIKCL_s1, SLTPPT_s1, SNDPPT_s1)) 
dim(soil)
```


Merge soil data with main file

```{r}
rice_rf1 <- merge (rice_rf, soil)
dim(rice_rf1)
summary(rice_rf1)

dim(rice_rf1)
ls(rice_rf1)
colnames(rice_rf1)
rice_rf1$SOC <- as.numeric(rice_rf1$SOC)
rice_rf1$PH <- as.numeric(rice_rf1$PH)
rice_rf1$CLYPPT_s1 <- as.numeric(rice_rf1$CLYPPT_s1)
rice_rf1$SLTPPT_s1 <- as.numeric(rice_rf1$SLTPPT_s1)
rice_rf1$SNDPPT_s1 <- as.numeric(rice_rf1$SNDPPT_s1)
rice_rf1$PHIKCL_s1 <- as.numeric(rice_rf1$PHIKCL_s1)
rice_rf1$q112_dist <- as.factor(rice_rf1$q112_dist)
dim(rice_rf1)
#write.csv(rice_rf1, "rice_rf1.csv")
```

Explore a bit with correlatioin plots
```{r}
#rice_c <- rice_rf1[, c(3,4,5,6,7,8,9,10,18,22,23,24,25,26,27,28,29,30,31,32,33,34,35,36,37,38,39,40,41)]
#View(rice_c)
#library(corrplot)
#dim(rice_c)
#M <- cor(rice_c)
#write.csv(M, "correlation.csv")
#corrplot(M, method="ellipse")
#corrplot(M, method="number")
#str(rice_c)
# I removed hhid, districts, lat & longitude colum
```

## k-means clustering 


```{r}
vars1 <- c("q117_gpslat", "q118_gpslong", "s_yield", "fym_kgha", "Nitrogen_kgha", "P2O5_kgha", "K2O_kgha", "Zn_kgha", "q244_noirrigrice","seeding_doy","transplant_doy","harvest_doi","June_15r","June_30r", "July_15r", "July_31r", "Aug_15r", "Aug_31r", "Sept_15r", "Sept_30r","SOC","PH", "CLYPPT_s1", "PHIKCL_s1", "SLTPPT_s1", "SNDPPT_s1")
rice_km <- rice_rf1[, vars1]  

##K-means clustering for wheat
#library(sp)
#isolates the covariates
set.seed(1983)
covar <- which(names(rice_km) == "q117_gpslat"):which(names(rice_km) == "SNDPPT_s1")
scaledR <- scale(rice_km[,covar])

#K-means clustering
wss <- (nrow(rice_km))*sum(apply(rice_km,2,var))
for (i in 1:26) wss[i] <- sum(kmeans(rice_km, 
                                     centers=i)$withinss)
plot(1:26, wss, type="b", xlab="Number of Clusters",
     ylab="Within groups sum of squares")


#Using the "elbow method" it appears that the data is clustering in 2 (slighltly tilted) 
rice_km$km3 <- kmeans(scaledR, centers=3, nstart=20, iter.max=20)$cluster
kmCentroids <- aggregate(list(rice_km[,covar]), by=list(cluster=rice_km$km3), FUN=mean)
k_meanstable <- rbind(t(table(rice_km$km3)), t(round(kmCentroids, 2)))

#write.csv(k_meanstable, "k_meanstable.csv")
```

### Ploting clusters
```{r, message=FALSE}
library(ggmap)
names(rice_km)
#qmplot(q118_gpslong, q117_gpslat, data=rice_km, color=as.factor(km3))
```


## For RandomForest model

```{r, message=FALSE}
#summary(rice_rf1)
#data=(rice_rf1)
vars2 <- c('s_yield', 'fym_kgha', 'Nitrogen_kgha', 'P2O5_kgha', 'K2O_kgha', 'Zn_kgha', 'q202_soiltype', 'q204_landtype', 'q205_varietytype', 'q2012_seedsource', 'q244_noirrigrice', 'q281_weedensity', 'q282_weedcontrol', 'transplant_doy', 'seeding_days', 'June_15r', 'June_30r', 'July_15r', 'July_31r', 'Aug_15r', 'Aug_31r', 'Sept_15r', 'Sept_30r', 'SOC', 'PH', 'CLYPPT_s1', 'SLTPPT_s1', 'SNDPPT_s1')
rice_rf1 <- rice_rf1[, vars2]
```

## Creating random forest model

```{r}
library(randomForest)
set.seed(1234567)
selected <- c(as.character(imp[1:27, 1]),'s_yield')
model1 <- randomForest(s_yield~.,data=train[,selected],replace=T,ntree=1000)
#model1
par(mfrow=c(1,2))
varImpPlot(model1,main='Variable Importance Plot: Final Model',pch=16,col='blue')
plot(model0, main='Error vs No. of trees plot: Final Model',col='blue')

#Base prediction (extra scripts lines from 126 to 132)..........(~!)
#pred1 <- predict(object=model0, newdata = train)
#actual1 <- train$s_yield
#result1<-data.frame(actual1=actual1, predicted=pred1)
#paste('Mean Squared error: ',mean(model0$mse))
#paste('Root Mean Squared error: ',mean(sqrt(model0$mse)))
#write.csv(result1, "rf_b.csv")
```


## Final solution 

```{r}
pred<-predict(object=model1,newdata=test)

actual<-test$s_yield
mf <-data.frame(actual=actual,predicted=pred)
paste('Mean Squared error: ',mean(model1$mse))
paste('Root Mean Squared error: ',mean(sqrt(model1$mse)))


#R-squre predicting from RF predicited vs observed pfpn
RF.rmse<-round(sqrt(mean( (mf$actual-mf$predicted)^2 , na.rm = TRUE )),0)
print(RF.rmse)

# R-square
RF.r2<-round(summary(lm(actual~predicted, mf))$r.squared,3)
print(RF.r2)
```


## Final solution 
plotting predicted Vs observed

```{r}
library(ggplot2)
ggplot(mf)+
  geom_point(aes(x=actual,y=predicted),alpha=0.6)+
  ggtitle('Random Forest "rice_yield"') +
  scale_x_continuous("Observed rice yield (kg/ha)",
                     limits=c(0,10000),
                     breaks=seq(0, 10000, 1000))+
  scale_y_continuous("Predicted rice yield (kg/ha)",
                     limits=c(0,10000),
                     breaks=seq(0, 10000, 1000)) +
  theme(axis.line = element_line(colour = "black"),
        axis.text.y=element_text(size=8,angle = 90, hjust = 0.5, vjust=1),
        axis.text.x = element_text(size=8)) +
  geom_abline(intercept=0,size = 0.5)+
  geom_smooth(aes(x=actual,y=predicted), method = "lm", se = FALSE, colour="red",linetype = 2,size = 0.75)+
  annotate("text", x=700, y=9000, label= paste("RMSE:",RF.rmse))+
  annotate("text", x=700, y=8500, label= paste("R^2: " ,RF.r2), parse=T)
paste('Function Call: ', model1$call)
```


## Partial dependence plots

```{r}
vars <- colnames(rice_rf1)
vars <- vars[vars != "s_yield"]
par(mfrow=c(2,2))
for (v in vars[1:4]) {
  partialPlot(model1, rice_rf1, c(v), ylab = "Yield", xlab=v, main="")
}

```


## CART all the rice-survey data 

```{r}
library(rpart)				     
library(rattle)					
library(rpart.plot)				
library(RColorBrewer)	

fit.all <- rpart(s_yield~., rice_rf1, method = "anova", parms =list("gini", loss=lmat, byrow=TRUE))
fit.all	

# Cross validation
printcp(fit.all) # display the results
plotcp(fit.all) # visualize cross-validation results
summary(fit.all) # detailed summary of splits

# Prune tree # trim nodes with a complexity of less than ?cp?
pfit.all<- prune(fit.all, cp= 0.014)
#fancyRpartPlot(pfit.all, cex=0.7, main=, sub, caption)    # Plot tree
fancyRpartPlot(pfit.all, cex=0.7)    # Plot tree

pfit.all
summary (pfit.all)	
```

