# Open Street Map data

A good source of roads data is Open Street Map (OSM). (expand what it is, and limitations (completeness?))

You can download OSM data in *R* using the [osmdata](https://cran.r-project.org/web/packages/osmdata/index.html) package. You can only download a certain amount of data in a single request, so typically you need to a number of requests for adjacent regions and then combine the results. We used the first level administrative boundaries of Tanzania (also downloaded with the geodata package) for this. 

```{r gadm, message=FALSE, warning=FALSE}
library(geodata)
adm <- gadm("TZA", level = 1, path=".")
```

We want data for primary, secondary and tertiary highways in Tanzania. Below is a helper function to accomplish this. 

```{r rdtp}
library(osmdata)

roadtypes <- c("primary", "secondary", "tertiary")

getOSMdata <- function(obj, key, value) {
    b <- sp::bbox(obj)
    q <- osmdata::opq(b)
    q <- osmdata::add_osm_feature(q, key, value)
    d <- osmdata::osmdata_sp(q)
    d$osm_lines[, key]
}

```

The below takes a long time to run. It may also fail because of the relative fraility of the OSM servers (or rather then enourmous number of requests that they get).

```{r getOSM, eval=FALSE}
x <- lapply(seq_along(adm), function(i) getOSMdata(adm[i, ], "highway", roadtypes))
```

Now put the results together 


```{r processOSM, eval=FALSE}
# combine the results for each area
rd <- do.call(bind, x)
# remove anything outside of Tanzania
rd <- crop(rd, adm)
# aggregate by type, also removing duplicates line parts
roads <- aggregate(rd, "highway")
```

The OSM data is very detailed. Much more than we need in most analysis. We simplified it with the `rmapshaper` package to store less data and so that the example runs faster. 

```{r simplr, eval=FALSE}
library(rmapshaper)
roads <- ms_simplify(roads, keep=0.01)
```

