---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Design


## Introduction

Now that we have a relationship between NDVI and mortality, we can design an insurance contract.

Load mortality predictions made the previous [section](5_mortality-model.html).

```{r d1}
d <- readRDS("pred_mort2.rds")
```

Remove some of the higher mortality predictions that occur with high ndvi. They are not relevant for us.

```{r d2}
d$predicted_mortality[d$NDVI > 0.1] <- 0.1
```

Let us find a *trigger* above which the insurance pays. We set it so that we expect the insurance to pay out once every five years. 

```{r d3}
trigger <- quantile(d$predicted_mortality, 0.8)
round(trigger, 2)
```

The trigger first occurs at this *zNDVI*

```{r d4}
tndvi <- max(d$NDVI[d$predicted_mortality > trigger])
round(tndvi, 2)
```

The payout --- in units of the monetary value of TLU --- can be computed as follows   

```{r d5}
d$pay <- pmax(0, d$predicted_mortality - trigger)

plot(d$NDVI, d$pay, cex=.1, xlab="NDVI", ylab="payout")
```

which is distributed over households, by years as follows

```{r d6}
boxplot(pay~year, data=d)
```

We can simplify the relationship between NDVI and mortality like this 

```{r d7}
nd <- d[d$NDVI < tndvi, ]
m <- lm(pay ~ NDVI, data=nd)
plot(pay ~ NDVI, data=nd, cex=.1)
abline(m, col="red")
cf <- coefficients(m)
cf
```

## Contracts

We can now design different contracts. The first would pay out according the model m above 

```{r dd8}
payout <- function(zNDVI) {
  pmax(0, -0.24 + -0.2675 * zNDVI)
}  

zndvi <- seq(-3, 2, .1)
plot(zndvi, payout(zndvi), type="l", col="red")
```

Another contract, folling Vrieling et al's early payout scheme, triggers payouts when *zNDVI* is less than *-0.842*.

```{r d8}
# need to adjust for payout scale (not TLUs?)
epay <- function(zNDVI, trigger=-0.842, exit=-2.326, frac=0.5) {
  zNDVI <- pmax(zNDVI, exit)
  frac * pmax( 0, (trigger-zNDVI) / (trigger-exit) )
}

plot(zndvi, epay(zndvi), type="l")
```

Plot the payout functions together.

```{r d10}
plot(zndvi, epay(zndvi), col="blue", type="l", lwd=2)
lines(zndvi, payout(zndvi), type="l", col="red", lwd=2, lty=2)
legend("topright", c("contract", "epay"), col= c("red", "blue"), lty=c(2,1), lwd=2)
```


