# Soil compaction effect on rice

## 1. Background and introduction

This case study shows how to do principal component analysis and stepwise multiple regression in R. 

Here we follow the analysis of the study by Singh et al (2017), entitled "Depth of soil compaction predominantly affects rice yield reduction by reproductive-stage drought at varietal screening sites in Bangladesh, India, and Nepal". The study was published in the journal Plant Soil. You can access the [article](https://link.springer.com/article/10.1007/s11104-017-3265-2) (it is behind a paywall) online. The data was also made available on Harvard Dataverse and can be [downloaded](https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/SNQZTT). 

--- Things to be noted here: Variables released in the dataset are not the same as in the paper. Results in the case study are different from the paper.---


## 2. Get the data

```{r}
library(agro)
ff <- get_data_from_uri("https://doi.org/10.7910/DVN/SNQZTT", ".")
ff
```


### 2.1. Import data

The data is provided as a tab file. This file can be read by read.table function in R.
```{r import table}
library(readxl)
# Import table of data
wdata <- read_excel(ff[1])

# Check column names to see the variable provided
colnames(wdata)

```


### 2.2. Subset data

Subset data for use and delete the information not use in the following analysis
```{r subset data}

wdata1 <- wdata[, !(colnames(wdata) %in% c("Year", "location", "Ave yield control", "Ave yield stress"))]

```


## 3. Explore data


### 3.1. Step-wise multiple regression for table 5 in the paper

Yield reduction by drought across three years of study as a function of rainfall and soil-related parameters as determined by step-wise multiple regression

3.1.1 First, use regsubsets(), which is a function in r for model selection
```{r}

# Make a multiple linear regression model with all explanatory variables and yield reduction as the response variable
# Load packages
library(car)

# Using the dot notation for shorter code, which means everything else in the data frame
model_full <- lm(`% yield reduction` ~ ., wdata1)

# Start with all variables and develop a good model using regsubsets() from the leaps package. Select the model with the minimum BIC

# Load packages
library(leaps)

# Using regsubset
model_regsub <- regsubsets(`% yield reduction` ~ ., data = wdata1)

# View results
summary(model_regsub)

# Visually determine best model by BIC (subset( ) are bic, cp, adjr2, and rss.)
summary(model_regsub)$bic
subsets(model_regsub, statistic = "bic", legend = F)

# Creat the best model
model_reduced <- lm(`% yield reduction` ~ `Rainfall 60-100 DAS` + `ave tensiom 60-100` + `max tensiom 60-100` + `aver water table 60-100 DAS` + `bd 5-10 cm` + pH + `% Clay` + `depth of max penetrom`, data = wdata1)

summary(model_reduced)

# Compare the best model with the full model 
anova(model_full, model_reduced)

```


3.1.2. Second, use stepAIC() function from MASS package perform step model selection by AIC including forward, backward and both
```{r}
library(MASS)

# Stepwise Regression
model_step <- stepAIC(model_full, direction = "both")

# Display results
model_step$anova

# Final model
model_final <- lm(`% yield reduction` ~ `Rainfall 60-100 DAS` + `ave tensiom 60-100` + `max tensiom 60-100` + `aver water table 60-100 DAS` + `bd 5-10 cm` + `bd 25-30 cm` + pH + `Avail-P ppm` + `Exch-K ppm` + `% Clay`, data = wdata1)

# Compare models (best reduced model from the previous analysis vs. the final model from this analysis)
anova(model_reduced, model_final)
```


### 3.2. Principal component analysis for table 6

Eigenvector values from a principal component analysis of yield reduction by drought a various soil characteristics at the research station drought screening sites characterized in this study.
```{r}

# Perform  a PCA on the 20 varaibles using the correlation matrix

# Load package
library(stats)

# Pricipal Component Analysis (When scale=TRUE, the correlation matrix is used, omit data wich is NA)
sr.pca <- prcomp(na.omit(wdata1), scale = TRUE)

# Display results
summary(sr.pca)

# Using a scree plot to check how many PC's to retain (find elbow)
plot(sr.pca, type = "l", main ="PCA of dataset")

# Check eigenvalues for each PC
summary(sr.pca)$sdev^2

# Check Eigenvector values for each PC's
re <- eigen(cor(na.omit(wdata1)))

```


### 3.3. Principal component analysis for figure 5

Biplot from a principal component analysis of yield reduction by drought a various soil characteristics at the research station drought screening sites characterized in this study. Parameter abbreviations are described in Table 5
```{r}

# Make a biplot
biplot(sr.pca)

```





