# Crop modelling with experimental data

## 1. Background and introduction

This case study is going to show how to apply crop model to simulate crop growth. WOFOST model and data from "Field experimental data for crop modeling of wheat growth response to nitrogen fertilizer, elevated CO2, water stress, and high temperature" were used. 

The data organized by Kassie et al (2018) was published in the Open Data Journal for Agricultural Research. You can freely access the [article](https://library.wur.nl/ojs/index.php/ODJAR/article/view/15828) published in the Open Data Journal for Agricultural Research (ODJAR). The data can be [downloaded](https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/V4P6PU) from the ODjAR Dataverse.
 
The data includes five experiments. Here we only take one study about water deficient experiment at Lincoln in New Zealand as an example, to see how to apply the WOFOST model in *R* to simulate crop growth under water limitations. 

## 2. Load R packages needed for this case study

```{r package}
library(Rwofost)
library(dplyr)
library(tidyr)
library(stringr)
```


## 3. Prepare the data

```{r getData}
library(agro)
ff <- get_data_from_uri("doi:10.7910/DVN/V4P6PU", ".")
length(ff)
head(ff)
```

### 3.1. Import the data

Here, use readLines() to read wth file for weather data and read.delim() to read text files for management and results.

```{r import}

# Import weather data: NEWZLAND.WTH
f <- ff[basename(ff) == "NEWZLAND.WTH"]
f
weather <- readLines(f)

# Import management data
f <- ff[basename(ff) == "Crop management_New Zealand.txt"]
management <- read.delim(f)

# Import reults from field experiment
f <- ff[basename(ff) == "Experimental data_New Zealand.txt"]
results <- read.delim(f)
```


### 3.2. Organize the data

3.2.1. Organize weather data for the model

The weather data is in character class, need to build a data frame for weather.

```{r weatherdata}

# Look at the weather data
head(weather)

# Remove the first 5 characters which is the information of this weather file 
w1 <- weather[-c(1:5)]

# Unlist characters
w2 <- unlist( strsplit( w1 , "\\ " ))

# Replace spaces with NA (as NA is filled as "99" in the orignal file)
w2[w2 == ""] <- NA

# Omit Na
w2 <- na.omit(w2)

# Make a matrix with 8 variables
w3 <- matrix(w2, ncol = 8, byrow = T)

# Formate date
date <- w3[ ,1]
year <- substr(date,1,2)
year1 <- paste0("19", year)
year <- as.integer(year1)
doy <- as.integer(substr(date,3,5))
date <- dateFromDoy(doy, year)

# Select the weather data used for the model
srad <- as.numeric(w3[ ,2]) * 1000
tmin <- as.numeric(w3[ ,4])
tmax <- as.numeric(w3[ ,3])
vapr <- as.numeric(rep(NA, nrow(w3)))
prec <- as.numeric(w3[ ,5])
wind <- as.numeric(rep(NA, nrow(w3)))
weather <- data.frame(date, srad, tmin, tmax, vapr, prec, wind)

```

3.2.2. Organize namagement data for the model

```{r management1}

# Subset irrigation info, here take treatment 11
irri <- management[ which (management$X.1 == "Irrigation" & management$X..Field.Data.site.information == "11"), ]

# Add column names
colnames(irri) <- as.character(unlist(management[143, ]))

```

3.2.3. Organize results data

```{r results}

# Read the observations of growth
f <- ff[basename(ff) == "Experimental data_New Zealand.txt"]
ob <- read.delim(f, skip = 15, header = TRUE)

# Subset results for treatment 11 
ob1 <- ob[which (ob$X..Definitions == "11"), ]

```


### 3.3. Download the weather data to replace NA.

```{r downloadweather}

# Downloade weather data from online by lon and lat to replace vapr and wind
we <- power_weather(172.39, -43.39)
wth <- we@data

# Subset period and replace the data
wth1 <- subset(wth, date >= as.Date('1991-01-01'))
weather$vapr <- wth1$vapr[1:nrow(weather)]
weather$wind <- wth1$wind[1:nrow(weather)]

```


## 4. Work with WOFOST model 

### 4.1 Input crop information

```{r crop}

# Choose wheat for this example
crop <- wofost_crop('wheat_fra')

```


### 4.2 Input soil information

```{r soil}

# Set soil type
soil <- wofost_soil('ec1')

```


### 4.3 Input management information

```{r management2}

# Set management condition
contrl <- wofost_control()

# Set location
contrl$latitude <- -43.39
  
# Set sowing date
contrl$modelstart <- as.Date("1991-06-08")

# Set irrigation dates and amount
contrl$IRRdates <- as.Date(irri[ ,5])
IRR <- as.numeric(as.character(irri[, 9]))
contrl$IRR <- matrix(IRR, nrow=1)

# Choose the production level (water limited condition)
contrl$IPRODL <- 1

```


### 4.4 Run the model and read output

```{r run}

# Run model and simulate wheat growth under water deficient condition
d <- wofost(crop, weather, soil, contrl)

# Read output results
#head(d)
#tail(d)

# Plot the grain yield vs. days after sowing
plot(d[,'step'], d[,'WSO'], type='l', xlab = "Days after sowing", ylab = "Dry grain yield (kg/ha)")

```

## 5. Compare the simulation results with field experimental observations

```{r compare}

# Read reults from field experiment
ob1$step <- as.Date(ob1$Observation.date) - contrl$modelstart + 1
colnames(ob1)[3] <- "WST"
colnames(ob1)[4] <- "LAI"
colnames(ob1)[5] <- "WSO"

ob1$WST <- as.numeric(as.character(ob1$WST))
ob1$LAI <- as.numeric(as.character(ob1$LAI))
ob1$WSO <- as.numeric(as.character(ob1$WSO))

# Compare results -- LAI 

# Plot the simulation results as a line
plot(d[,'step'], d[,'LAI'], type='l',xlab='Days',
     ylab='LAI', ylim = c(0,7), xlim = c(0,232))

# Add the observations as points
points (ob1[,'step'], ob1[, 'LAI'], col = "red")

# Add legend
legend("topleft", legend = c("Simulated", "Observed"), pch = c(NA, 1), lty = c(1, 0), col = c("black", "red"))


# Compare results -- Dry biomass
plot(d[,'step'], d[,'WST'], type='l',xlab='Days',
     ylab='Biomass (kg/ha)', xlim = c(0,232), ylim = c(0,16000))
points (ob1[,'step'], ob1[, 'WST'], col = "red")
legend("topleft", legend = c("Simulated", "Observed"), pch = c(NA, 1), lty = c(1, 0), col = c("black", "red"))

# Compare results -- Dry grain yield
plot(d[,'step'], d[,'WSO'], type='l',xlab='Days',
     ylab='Grain yield (kg/ha)', xlim = c(0,232), ylim = c(0, 3600))
points(ob1[, 'step'], ob1[,'WSO'], col = "red")
legend("topleft", legend = c("Simulated", "Observed"), pch = c(NA, 1), lty = c(1, 0), col = c("black", "red"))


#--- Simulation Results needs to be improved---# 
```

