---
output: html_document
editor_options: 
  chunk_output_type: console
---
# The QUEFTS model

## Introduction

This chapter describes the `QUEFTS` model. QUEFTS stands for *QUantitative Evaluation of the Fertility of Tropical Soils*. It is a rule-based model that can be used to estimate crop yield from soil properties, the amount of fertilizer applied, and an estimate of the yield that could be obtained when soil nutrients are in ample supply. It can also be used to estimate the amount of fertilizer needed to reach a particular yield. QUEFTS was first described by [Janssen et al., 1990](https://www.sciencedirect.com/science/article/pii/001670619090021Z). See [Sattari et al., 2014](https://www.sciencedirect.com/science/article/abs/pii/S0378429013004097) for an evaluation and updates of the model.

To run QUEFTS for a particular location, you need data on local soil quality, and we discuss these first. You also need crop parameters and the location specific attainable yield. In this context, attainable yield is the yield that would be reached if crop growth is not limited by low levels the three macro-nutrients nitrogen (N), phosphoros (P) and potassium (K).

## Chapter requirements

We use *R* package `Rquefts`. You can [install](/installation.html) it from CRAN.

## Native soil supply of nutrients

The first step in using a QUEFTS model is to estimate the native soil supply of the three macro-nutrients N, P, K. Other nutrients are ignored here, although the model could be extended to include them. 

The native supply of nutrients is the amount that is availble to a crop when no fertilizer is applied to the soil. The native nutrient supply is estimated from chemical properties of the topsoil. 

The equations used by QUEFTS to calculate the native supply of nutrients from soil properties were derived from empirical data of field trials in Suriname and Kenya. They were thought to be applicable to well drained, deep soils, that have a pH(H20) in the range 4.5-7.0, and values for organic carbon below 70 g/kg, P-Olsen below 30 mg/kg and exchangeable potassium below 30 mmol/kg [Janssen et al., 1990](https://www.sciencedirect.com/science/article/pii/001670619090021Z). [Sattari et al., 2014](https://www.sciencedirect.com/science/article/abs/pii/S0378429013004097) expanded the equations to be applicable to a wider range of conditions, and we will use these improved equations here. 

Below we compute the nutrient supply for a location with an average temperature of 25C, a pH of 6, the organic C concentration is 10 g/kg, and exchangeable potassium is 1 mmol/kg. Phosphorus extracted with the Olsen method is 10 mg/kg; and the total amount of P is is 500 mg/kg.

```{r quefts_supply_1} 
library(Rquefts)
s <- nutSupply2(temp=25, pH=6, SOC=15, Kex=1, Polsen=10, Ptotal=500)
s
```


We see that the soils supply of N is `r round(s[1,1])` kg/ha, that the P supply is `r round(s[1,2])` kg/ha, and the K supply is `r round(s[1,3])` kg/ha. 

Now let's compute nutrient supply across a range of pH values, keeping the other values constant.


```{r quefts_supply_10, fig.with=8} 
ph <- seq(from=4.5, to=8.5, by=0.1)
s <- nutSupply2(temp=25, pH=ph, SOC=15, Kex=1, Polsen=10, Ptotal=500)
phs <- cbind(ph, s)
head(phs)
par(mfrow=c(1,3))
plot(phs[, c(1,2)])
plot(phs[, c(1,3)])
plot(phs[, c(1,4)])
```

We see that the pH affters the supply of nutruents in different ways for N, P, and K. The P-supply is highest at a *pH*=6 -- reflecting the fact that phosphoros becomes less available to plants at low or high pH. N-supply decreases with *pH* (everything else, such as the amount of soil carbon,  being equal), and K-supply increases with *pH* (given the amount of exchangeble K).

Let's do this once, more but with different levels of some of the other properties.

```{r quefts_phsupply} 
s1 <- nutSupply2(25, ph, SOC=15, Kex=5,  Polsen=5, Ptotal=5*55)
s2 <- nutSupply2(25, ph, SOC=30, Kex=10, Polsen=10, Ptotal=10*55)
s3 <- nutSupply2(25, ph, SOC=45, Kex=20, Polsen=20, Ptotal=20*55)
```

We can use the results to make plots of the effect of pH on N, P and K supply, for these three situations (low, medium and high levels of organic matter, exchangeble K, and Olsen-P).

```{r quefts_phsupply2, fig.width=10} 
par(mfrow=c(1,3))
sup <- c("N_base_supply", "P_base_supply", "K_base_supply")
for (i in 1:3) {
  yb <- paste(sup[i],"(kg/ha)")
  ym <- c(0, max(s3[, sup[i]]))
  plot(ph, s1[,sup[i]], xlab="pH", ylab=yb, type="l", ylim=ym, lty=1, las=2)
  lines(ph, s2[, sup[i]], lty=2)
  lines(ph, s3[, sup[i]], lty=3)
}
legend("bottomleft", paste("treatment", 1:3), lty=1:3)
```


Now let's look at the nutrient supply for differnt levels of soil Carbon.

```{r quefts_omsupply, fig.width=10} 
# unit is g/kg
om <- seq(0, 70, 5)
# three combinations
s1 <- nutSupply2(temp=25, pH=5.5, SOC=om, Kex=5, Polsen=5, Ptotal=5*55)
s2 <- nutSupply2(temp=25, pH=5.5, SOC=om, Kex=10, Polsen=10, Ptotal=10*55)
s3 <- nutSupply2(temp=25, pH=5.5, SOC=om, Kex=20, Polsen=20, Ptotal=20*55)

# and plot
par(mfrow=c(1,3))
for (supply in c("N_base_supply", "P_base_supply", "K_base_supply")) {
  plot(om, s3[,supply], xlab="Soil Carbon (g/kg)", ylab=paste(supply, "(kg/ha)"), type="l", ylim=c(0,max(s3[,supply])), lty=1)
  lines(om, s2[,supply], lty=2)
  lines(om, s1[,supply], lty=3)
}
```

This shows that organic carbon has a strong positive effect on available N and P, but a negative effect on K availability (given the amount of exchangable K). 

To use QUEFTS we need to assume that these relationships are robust --- that is, not too far off. But with experimental data, these relationships could be re-estimated and improved. 


## Nutrient uptake and production

With the native soil nutrient supply, and possibly supply from fertilizer, QUEFTS computes the nutrient uptake by a crop, and from that it estimates crop yield. 

Uptake of each nutrient is estimated from the soil supply of that nutrient, the supply of the other two nutrients, and the maximum nutrient concentrations in the vegetative and generative organs of the crop. These maximum concentrations are crop-specfic parameters that we will use below when we run the model.

The estimated uptake of N, P, and K is then used to estimated biomass production ranges for each nutrient. For each pair of nutrients two yield estimates are calculated. For example, production for the N uptake is dependent on the production range for the P and the for the K uptake. This leads to six combinations describing the uptake of one nutrient given maximum dilution or accumulation of another. The nutrient-limited production is then estimated by averaging these six estimates for paired nutrients. An estimate based on two nutrients may not exceed the upper limit of the yield range of the third nutrient, that is, the concentration of a nutrient cannot be lower than its maximum dilution level. Neither may the production estimates exceed the specified attainable (maximum) production. 


## Create and run a model

### Quick start

To run QUEFTS, the first step is to create a model. You can create a model with default parameters like this

```{r quefts_100} 
library(Rquefts)
q <- quefts()
```

And then run the model

```{r quefts_110} 
run(q)
```

Normally you would initialize a model with parameters of your choice. 

### Soil parameters

```{r quefts_120soil} 
soil <- quefts_soil()
class(soil)
str(soil)
soil$UptakeAdjust
```

We see that there are 7 soil parameters (one if which is a matrix). The first three give the native soil supply, as we discussed above. The "recovery" parameters set the fraction of avaiable nutrients that a crop may take up (the remainder is either lost or chemically or biologically captured). The UptakeAdjst matrix serves to adjust the nutrient availability given the length of the growing season. The standard season length is 120 days. The longer the season, the higher the total nutrient supply.


### Crop parameters

The Rquefts package comes with a set of standard parameters for a number of crops. We get a list of them if we request a crop that does not exist.

```{r quefts_120crops, error=TRUE} 
quefts_crop(name="x")
```

We'll use barley
```{r quefts_120bar, error=TRUE} 
crop <- quefts_crop(name="Barley")
str(crop)
```

Again, we have a list of paramters. These are the mostly the minimum and maximum delution parameters, as described above, for the vegetative and generative (storage) organs, and for N, P, and K. 

If the crop has biological nitrogen fixation (in symbiosis with *Rhizobium*) you can use the Nfix parameter to determine the fraction of the total nitrogen supply that is supplied by biological fixation. Note that Nfix is a constant, while in reality nitrogen fixation is amongst other things dependent on the amount of mineral nitrogen available in the soil.


### Other parameters

You can set the fertilizer applied with a list like this

```{r quefts_120fert} 
fertilizer <- list(N=50, P=0, K=0)
```

We set the attainable biomass accumlation to 2200 kg/ha for leaves, 2700 kg/ha for stems, and 4800 kg/ha for grain. All these numbers are expressed as dry matter weigts. As a rule of the thumb, for cereals, you can assume a harvest index of 0.5. That is, the grain yield is about half the biomass. The other half is divided by the leaves (45%) and stems (55%). We set the seasonlength to 110 days.

```{r quefts_120fertbiom} 
biomass <- list(leaf_att=2200, stem_att=2700, store_att=4800, SeasonLength=110)
```

### Create model

And we can create a model with these parameters. 
```{r quefts120mod} 
q <- quefts(soil, crop, fertilizer, biomass)
```

You can inspect the model with `str` (short for "structure"):
```{r quefts120120} 
str(q)
```

And you can inspect and change parameters. 


```{r quefts_120dollar} 
q$N
q$N <- 100
q$N
```

### model output

And, of course, you can run the model.

```{r quefts_120run} 
output <- run(q)
round(output, 1)
```


The results show the nutrient limited production of leaves, stems, and the storage organ, as well as the N, P, and K supply and uptake. The N, P, and K gap show the amounts of fertilizer that would be required to reach the attainable yield.  


## Explore 

Let's have a look how rice yield is affected by N fertilization. First set up some parameters
```{r quefts_130}
rice <- quefts_crop("Rice")
q <- quefts(crop=rice)
q$leaf_att <- 2651
q$stem_att <- 5053
q$store_att <- 8208
fert(q) <- list(N=0, P=0, K=0)
N <- seq(1, 200, 10)
```

Now set up an output matrix, and run the model for each fertlizer application. 

```{r quefts_140}
results <- matrix(nrow=length(N), ncol=12)
colnames(results) <- names(run(q))
for (i in 1:length(N)) {
	q["N"] <- N[i]
	results[i,] <- run(q)
}
yield <- results[,"store_lim"]
yield
```

And make a plot

```{r quefts_150}
plot(N, yield, type="l", lwd=2)
```

We can also look at interactions by changing two nutrients. Here we change N and P fertilization levels, while keeping K constant (either 0 or 200 mmol/kg)

```{r quefts_160}
f <- expand.grid(N=seq(0,400,10), P=seq(0,400,10), K=c(0,200))
x <- rep(NA, nrow(f))
for (i in 1:nrow(f)) {0
	q$N <- f$N[i] 
	q$P <- f$P[i] 
	q$K <- f$K[i] 
	x[i] <- run(q)["store_lim"]
}
x <- cbind(f, x)
head(x)
```

To display the results we can treat the values as a raster

```{r quefts_plotraster, fig.width=10}
library(raster)
r0 <- rasterFromXYZ(x[x$K==0, -3])
r200 <- rasterFromXYZ(x[x$K==200, -3])

par(mfrow=c(1,2))
plot(r0, xlab="N fertilizer", ylab="P fertilizer", las=1, main="K=0")
contour(r0, add=TRUE)
plot(r200, xlab="N fertilizer", ylab="P fertilizer", las=1, main="K=200")
contour(r200, add=TRUE)
```


## Callibration

If you have your own experimental data, you could callibrate QUEFTS. Here I illustrate some basic approaches that can be used for callibration.First I generate some data. 

```{r quefts_180} 
set.seed(99)
yldfun <- function(N, noise=TRUE) { 1000 +  500* log(N+1)/3 + noise * runif(length(N), -500, 500) }
N <- seq(0,300,25)
Y <- replicate(10, yldfun(N))
obs <- cbind(N, Y)
```

We will use Root Mean Square Error (RMSE) to assess model fit. 

```{r quefts_190} 
RMSE <- function(obs, pred) sqrt(mean((obs - pred)^2))
```

Let's see how good the default parameters work for us.

```{r quefts_200} 
q <- quefts()
q$P <- 0
q$K <- 0

results <- matrix(nrow=length(N), ncol=12)
colnames(results) <- names(run(q))
for (i in 1:length(N)) {
	q$N <- N[i]
	results[i,] <- run(q)
}

yield <- results[,'store_lim']
RMSE(obs[,-1], yield)
```

The RMSE is quite high. 

We create a function to minimize with the optimizer. Here I try to improve the model by altering four parameters.

```{r quefts_210} 
vars <- c('soil$N_base_supply', 'soil$N_recovery', 'crop$NminStore', 'crop$NmaxStore')
f <- function(p) {
	if (any (p < 0)) return(Inf)
	if (p['crop$NminStore'] >= p['crop$NmaxStore']) return(Inf)
	if (p['soil$N_recovery'] > .6 | p['soil$N_recovery'] < .3) return(Inf)
	q[names(p)] <- p
	pred <- run(q)
	RMSE(obs[,-1], pred['store_lim'])
}

# try the function with some initial values
x <- c(50, 0.5,  0.011, 0.035)
names(x) <- vars
f(x)
```

Now tht we have a working functon, we can use optimization methods to try to find the combination of parameter values that is the best fit to our data. 

```{r quefts_220} 
par <- optim(x, f)
# RMSE
par$value
# optimal parameters
par$par
```


