# Spatial QUEFTS

Sebastian Palmas and Robert Hijmans

## Introduction 

In this chapter we show how to run QUEFTS with spatial data to make predictions. See the [previous](quefts.html) chapter for a description of the model itself. 

## Chapter requirements

We use the *R* packages `Rquefts`, `terra` and `reagro`. See [these instructions](/installation.html) about installing these packages.

## Soil data

We use soil raster data for Tanzania from [SoilGrids](https://soilgrids.org/). We changed the pH values from the original data source by dividing them by 10 to get the correct numbers.

```{r Loading soil rfiles}

library(agrodata)
SOC <- reagro_data("TZA_ORC")
Kex <- reagro_data("TZA_EXK")
pH <- reagro_data("TZA_PH")
```

Need to fix this, there is P data in soilgrids:

"There is no P olsen raster in soilgrids, in the code below we create a constant P Olsen layer by using the soilC layer as a template and assigning all cells a value of 15."


From these chemical soil properties, we can use QUEFTS to compute the soil nutrient supply.

```{r compute soil nutrient supply}
library(Rquefts)
supply <- rast(pH, nlyr=3)
values(supply) <- nutSupply1(pH, SOC, Kex, Polsen=15)
supply
```

We have some negative values (let's remove them. Let's see what happens when we use real P values)

```{r clamp}
supply <- clamp(supply, 1, Inf)
names(supply) <- c("Ns", "Ps", "Ks")
supply
plot(supply)
```


## Water limited yield

For attainable yield, we use a raster of water-limited yield estimated by the GYGA project. 

```{r WY}
Ya <- reagro_data("TZA_YW")
```

## Running QUEFTS

```{r set up a q model}
library(Rquefts)
maize <- quefts_crop("Maize")
fertilizer <- list(N=64, P=20, K=0)
q <- quefts(crop=maize, fert=fertilizer)
```

In this example, 64 kg/ha of N and 20 kg/ha of P comes from using 100 kg/ha of Urea and 100 kg/ha of DAP.

```{r Running}
yield <- rast(Ya)
values(yield) <- run(q, supply, Ya)
plot(yield)
```

