# Data bases

## Introduction 

## Chapter requirements

We use *R* package `agrodata`. See these [installation](/installation.html) instructions.

## Sacks data

There are data for 20 crops, some are divided into two groups (winter / spring) or (main / 2nd season).

```{r sacks1, message=FALSE}
library(agrodata)
agrodata::crop_calendar_sacks()
```

We choose "Maize (main season)"

```{r sacks15, message=FALSE}
r <- agrodata::crop_calendar_sacks("Maize (main season)")
r
```

Note that there are 13 variables in this dataset for each crop.

```{r sacks20}
names(r)
```

The mean day of harvest 

```{r sacks30, fig.width=8, fig.height=5}
plot(r[["harvest"]])
```

But harvesting is not all done on one day. It is estimated to start on this day:

```{r sacks35, fig.width=8, fig.height=5}
plot(r["harvest.start"])
```
And end on

```{r sacks40, fig.width=8, fig.height=5}
plot(r["harvest.end"])
```

"harvest.range" is the difference between "harvest.start" and "harvest.end"

The "harvest.*" variables are also avaiable for planting.

```{r sacks45, fig.width=8, fig.height=5}
plot(r["harvest.end"])
```

"tot.days" is the number of days between planting and harvest. 

```{r sacks50, fig.width=8, fig.height=5}
plot(r["tot.days"])
```

"harvested.area" has the crop area in ha, and "harvested.area.fraction" has it as a fraction of the grid cell. 

```{r sacks60, fig.width=8, fig.height=5}
plot(r["harvested.area.fraction"])
```
