---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Ecocrop

```{r, include=FALSE}
library(terra)
```

## Introduction 


## Chapter requirements

We use *R* packages `ecocrop` and `geodata`. See these [installation](/installation.html) instructions.


## Ecocrop 


## Run the model 

As in the previous chapter, we use climate data for Senegal from [WorldClim](/data/spatial/climate.html) data to find . 

```{r ecocrop100, message=FALSE}
library(geodata)
rain <- geodata::worldclim_country("Senegal", var="prec", path=".")
tavg <- geodata::worldclim_country("Senegal", var="tavg", path=".")

adm <- geodata::gadm("Senegal", level=1, path=".")
rain <- mask(rain, adm)
tavg <- mask(tavg, adm)
```

Create an ecocrop model for maize

```{r ecocrop110}
library(Recocrop)
crop <- Recocrop::ecocropPars("maize")
m <- Recocrop::ecocrop(crop)
```

Use the model to make predictions for Senegal

```{r ecocrop120}
plant <- predict(m, prec=rain, tavg=tavg)
p <- classify(plant > 0, cbind(0,NA)) * 1:12
pm <- median(p, na.rm=TRUE)
hv <- pm + m$duration
hv <- ifel(hv > 12, hv - 12, hv)
plot(hv)
lines(adm)
```

Or like this

```{r ecocrop150}
control(m, which_max=TRUE)
mplant <- predict(m, prec=rain, tavg=tavg)
mhv <- mplant + m$duration
mhv <- ifel(mhv > 12, mhv - 12, mhv)
plot(mhv)
lines(adm)
```
