# Cereal yield in West Africa

## Introduction

This case study follows the paper entitled "Cereal yield response to conservation agriculture practices in drylands of West Africa: A quantitative synthesis" by Bayala *et al*. (2012) which was published in the Journal of Arid Environments. You can access the [article here](https://www.sciencedirect.com/science/article/pii/S0140196311003120) (it is behind a paywall). The data was made available on the World Agroforestry Center [dataverse](https://doi.org/10.34725/DVN/1BLSQY).

This study provides a quantitative synthesis of the effect of conservation agriculture practices on crop yield in the drylands of West Africa. 


## Get the data

We can download the data with `get_data_from_uri` from the `agro` package.

```{r}
ff <- agro::get_data_from_uri("doi:10.34725/DVN/1BLSQY", ".")
ff
```

You can see that we have a single Excel file that we can read with `readxl::read_excel`.

```{r}
readxl::excel_sheets(ff)
``` 

There are five sheets in the file. We need the sheets that have the grain and straw data. 

```{r}
grain <- readxl::read_excel(ff, sheet = "Grain")
straw <- readxl::read_excel(ff, sheet = "Straw")
dim(straw)
dim(grain)
colnames(grain)
```

We wanted to quickly look at the "Rawdata" sheet too, to understand the difference. 

```{r}
raw <- readxl::read_excel(ff, sheet = "Rawdata")
dim(raw)
dim(straw) + dim(grain)
table(raw$Component)
```

It appears that records in `Rawdata` that are not straw or grain have been ignored. Creating multiple tables with subsets of the data is generally a bad idea because it may lead to errors due to copying, or when an error is corrected in one of the tables but not everywhere. We could have created the grain and straw subsets like this (showing for grain only)

```{r}
grain2 <- raw[tolower(raw$Component) == "grain", ]
``` 

To test if this is indeed gives us the same values, we first sort the both datasets. 

```{r}
grain  <- grain[order(grain$PubN, grain$Treatment), ]
grain2 <- grain2[order(grain2$PubN, grain2$Treatment), ]
```

And then we can compare them

```{r}
all(grain == grain2, na.rm=TRUE)
```

*Quod erat demonstrandum*


## Organize data

We also need to compute some new values (to be used below) and fix some values in the `Technology` variable in both tables.

```{r}
compute_and_fixme <- function(x) {

  x$D <- x$`Treatment yield` - x$`Control yield` 
  
  x$productivity = cut(x$`Control yield`, c(0, 0.5, 1, 1.5, 2, Inf))

  x$Technology[x$Technology == "Coppiced G. manure"] <- "Coppicing"
  x$Technology[x$Technology == "Rot-Ass"] <- "Rotations"
  x$Technology[x$Technology == "SWC"] <- "Soil water"
  x
}

grain <- compute_and_fixme(grain)
straw <- compute_and_fixme(straw)
```


## Tables 1 and 2


Now we can reproduce Table 1 from the article: *Summary of statistics of mean difference in grain yield (D, t/ha) across all crops in different practices of conservation agriculture in four West African Sahelian countries (Burkina Faso, Mali, Niger and Senegal).*

Now compute the summary statistics. We use a function, so that we can use the same code for the next table.

```{r tab1B}
make_table <- function(x) {
  # unique number of publications by technology
  N <- tapply(x$PubN, x$Technology, function(i) length(unique(i)))
  # observations by technology
  k <- table(x$Technology)
  # mean yield gain by technology
  mean <- tapply(x$D, x$Technology, mean)
  # quartiles of yield gain by technology
  q <- tapply(x$D, x$Technology, function(i) quantile(i, c(.75, .5, .25)))
  q <- do.call(cbind, q)
  iqr <- q[1,] - q[3,]
  ## count cases within classes
  # classify
  ycls <- cut(x$D, c(-Inf, 0, 1, 2, Inf))
  # tabulate
  z <- table(ycls, x$Technology)
  # compute percentage. Note that we need to transpose to divide
  # correctly, and then transpose back
  z <- 100 * t(t(z) / colSums(z))
  
  ## combine results
  tab <- rbind(N, k, mean, q, iqr, z)
  tab <- round(tab, 2)
  ## ordering
  tab <- tab[, c("Parkland", "Coppicing", "Green manure", "Mulching", "Rotations", "Soil water")]
  tab
}
```

Now use the function

```{r tab1C}
tab1 <- make_table(grain)
tab1
```

Results look very similar. Except for the classes counts. But the published table is odd. There are intervals missing, and it is unclear what the exact intervals are. 

Recreating Table 2, *Summary of statistics of mean difference in straw biomass yield (D, t/ha) in different practices of conservation agriculture in four West African Sahelian countries (Burkina Faso, Mali, Niger and Senegal)* is now straightforward, because we can re-use the function defined above for Table 1.


```{r tab2A}
tab2 <- make_table(straw)
tab2
```

Results are similar. But note, for example, that we find k=28 for Parkland, whereas the paper reports 26. 

## Figure 1. 

Figure 1 shows *Variation in mean difference in yield with management practices in four West African Sahelian countries (Burkina Faso, Mali, Niger and Senegal). Vertical bars indicate 95% confidence bands.*

```{r}
# Check the crop types
sort(unique(grain$Crop))
```

We need to create a column with "clean" crop names. I left "Millet-sorghum" observations out, as we do not know if it is millet or sorghum.

```{r}
### Select data for sorghum
grain$cropname <- ""
grain$cropname[substr(grain$Crop, 1, 7) == "Sorghum"] <- "Sorghum"
grain$cropname[grain$Crop %in% c("White sorghum", "Red sorghum")] <- "Sorghum"
grain$cropname[substr(grain$Crop, 1, 5) == "Maize"] <- "Maize"
grain$cropname[grain$Crop == "Millet"] <- "Millet"

# check by making a table
tab <- table(grain$Crop, grain$cropname)
head(tab, 15)
```


Now aggregate for the three crops of interest 
```{r}
par(mfrow=c(1,3))
for (crop in c("Sorghum", "Millet", "Maize")) {
    plotmeans(D~Technology, data=grain, subset=cropname==crop, connect=FALSE, 
          xlab= "", ylab="Grain yield", las=2, ylim=c(-1, 1.6))
    text(1, 1.5, crop)
}

```


## Figs 2 and 3 

As figure 2 and 3 are similar, Here figure 3 is replicated. *Variation in mean difference in yield (across cereal crops) with soil and water conservation practices and site productivity in four West African Sahelian countries (Burkina Faso, Mali, Niger and Senegal). The broken line represents the point where the treatment and control yields are equal. 1 and 2 defined low potential, 3 and 4 medium potential and above 4 defined high potential sites. Vertical bars represent standard errors.*

```{r}
g <- grain[grain$cropname != "", ]
g$productivity <- as.integer(g$productivity)
sciplot::lineplot.CI(productivity, D, group=cropname, data=g, 
          ci.fun=function(x) c(mean(x)-1.96 * sciplot::se(x), mean(x)+1.96 * sciplot::se(x)),
          xlab= "site productivity", ylab="Mean difference in yield", lwd=2)
abline(h=0, lwd=2, col="gray")

```


-----------------------------RMEL MODEL NOT SOLVED----------------------------------

### 3.5. Table 3. Effect of rainfall on cereal grain yield (across crops) response to conservation agriculture practices in four West African Sahelian countries (Burkina Faso, Mali, Niger and Senegal).

```{r}

# Seperate data for grain yield of parkland based on rainfall class

grain_pt1 <- grain_pt[which(grain_pt$Rainfall < 601), ]
grain_pt2 <- grain_pt[which(grain_pt$Rainfall >= 601 & grain_pt$Rainfall <= 800), ]
grain_pt3 <- grain_pt[which(grain_pt$Rainfall > 800), ]

library(lme4)
#lm_pt <- lmer(`Yield Gap` ~ Rainfall + (Rainfall|Site), data = grain_pt1)

### Not sure which is random effects of the model

```

