# Crop parameters

## Introduction 

The crop parameters used are fundamental to using WOFOST. They can be difficult to navigate because there are so many of them (48 in the current version). Here we present them all, so you can evaluate which parameters are of most relevance to you. Unless you are a specialist, you should use default values for most parameters for a crop of interest. However, you could consider using different values for the phenology parameters.  

## List of crops

Here is a list of the default crop parameter sets that come with Rwofost

```{r wofost_crop1}
library(Rwofost)
crops <- wofost_crop()
cat(crops, "\n")
```

You can get parameters for a single crop (variety) by like this. 

```{r wofost_crop10}
crop <- wofost_crop("barley")
```

You can also provide a filename to `wofost_crop` to read parameters from a file you created yourself. Have a look at the files in the Rwofost package. On your computer they are installed on what is returned by *R* if you run this: `system.file("wofost/crop", package="Rwofost")`

The object `crop` is a `list` with `r length(crop)` elements (some of the parameters are used to compute nutrient limitation but that has not been implemented yet --- so these can be ignored for now). 

```{r wofost_crop12}
class(crop)
length(crop)
str(crop[1:5])
```

Most are single numbers, but there are also two-row matrices (referred to as tables in the context of WOFOST). The names of these matrix-parameters end on TB (for table). Here is an example. 

```{r wofost_crop14}
crop$FOTB
```

You can get some (but not very informative) meta-data about the parameters by using the argument `describe=TRUE`.

```{r wofost_crop18}
crop <- wofost_crop("barley", describe=TRUE)
```


## Parameters
	
### Emergence

These parameters help determine how long it takes for the crop to emerge. This is only relevant when the  simulation of the crop starts at sowing, and not at emergence. For emergence to occur, a thermal time `TSUMEM` has to be reached, but summing up the daily difference in between the average temperature and `TBASEM` --- if the temperature is > `TBASEM`; and by capping the average temperature at `TEFFMX`.



```{r wofost_crop_20, echo=FALSE}
params <- read.csv("croppars.csv", stringsAsFactors=FALSE)
params[,4] <- gsub("-1", "^-1^", params[,4])
params[,4] <- gsub("2", "^2^", params[,4])
p <- params[params[,1] == "emergence", -1]
rownames(p) <- NULL
knitr::kable(p)
```

The default parameters for barley are: 

```{r wofost_crop_22}
crop$TSUMEM
crop$TEFFMX
crop$TBASEM
```

### Initial conditions

```{r wofost_crop_40, echo=FALSE}
p <- params[params[,1] == "initial", -1]
rownames(p) <- NULL
knitr::kable(p)
```

### Phenology

Phenology parameters are very important as they determine the "maturity type" (early, late) of the variety, and hence the length of the growing season. This can have a very strong effect of crop yield potential, but also on the risk to encounter drought stress. You can easily change the maturity type by changing the values for `TSUM1` and `TSUM2`

```{r wofost_crop_30, echo=FALSE}
p <- params[params[,1] == "phenology", -1]
rownames(p) <- NULL
knitr::kable(p)
```

Other WOFOST implementations may also have `DVSI`, the initial development stage (DVS), and `DVSEND`	development stage at harvest; but Rwofost does not have these.


### Green area

```{r wofost_crop_50, echo=FALSE}
p <- params[params[,1] == "green area", -1]
rownames(p) <- NULL
knitr::kable(p)
```

### Assimilation

```{r wofost_crop_60, echo=FALSE}
p <- params[params[,1] == "assimilation", -1]
rownames(p) <- NULL
knitr::kable(p)
```

### Conversion of assimilates into biomass

```{r wofost_crop_70, echo=FALSE}
p <- params[params[,1] == "conversion of assimilates into biomass", -1]
rownames(p) <- NULL
knitr::kable(p)
```


### Maintenance respiration

```{r wofost_crop_80, echo=FALSE}
p <- params[params[,1] == "maintenance respiration", -1]
rownames(p) <- NULL
knitr::kable(p)
```

### Partitioning of biomass

```{r wofost_crop_90, echo=FALSE}
p <- params[params[,1] == "partitioning", -1]
rownames(p) <- NULL
knitr::kable(p)
```

### Death rates

```{r wofost_crop_100, echo=FALSE}
p <- params[params[,1] == "death rates", -1]
rownames(p) <- NULL
knitr::kable(p)
```

### Water use

```{r wofost_crop_110, echo=FALSE}
p <- params[params[,1] == "water use", -1]
rownames(p) <- NULL
knitr::kable(p)
```

### Roots

```{r wofost_crop_120, echo=FALSE}
p <- params[params[,1] == "rooting", -1]
rownames(p) <- NULL
knitr::kable(p)
```

### CO~2~

```{r wofost_crop_130, echo=FALSE}
p <- params[params[,1] == "CO2", -1]
rownames(p) <- NULL
knitr::kable(p)
```

