# Crop parameters

## Introduction 


## Parameter files 

Available crops with Rwofost

```{r wofost_crop1}
library(Rwofost)
crops <- wofost_crop()
cat(crops)
```

Get parameters for a single crop

```{r wofost_crop10}
crop <- wofost_crop("barley")
str(crop[1:5])
```

The object `crop` is a `list` with `r length(crop)` elements. Most are single numbers, but there are also two-row matrices (referred to as tables in the context of WOFOST).



```{r wofost_crop_table, echo=FALSE}
params <- read.csv("croppars.csv", stringsAsFactors=FALSE)
rownames(params) <- NULL
```

## Parameter definitions

The crop parameters
	
### emergence

```{r wofost_crop_20, echo=FALSE}
p <- params[params[,1] == "emergence", -1]
rownames(p) <- NULL
knitr::kable(p)
```

### initial conditions

```{r wofost_crop_40, echo=FALSE}
p <- params[params[,1] == "initial", -1]
rownames(p) <- NULL
knitr::kable(p)
```


### phenology


```{r wofost_crop_30, echo=FALSE}
p <- params[params[,1] == "phenology", -1]
rownames(p) <- NULL
knitr::kable(p)
```

Other WOFOST implementations may also have `DVSI`, the initial development stage (DVS), and `DVSEND`	development stage at harvest; but Rwofost does not have these.


### green area

```{r wofost_crop_50, echo=FALSE}
p <- params[params[,1] == "green area", -1]
rownames(p) <- NULL
knitr::kable(p)
```

### assimilation

```{r wofost_crop_60, echo=FALSE}
p <- params[params[,1] == "assimilation", -1]
rownames(p) <- NULL
knitr::kable(p)
```

### conversion of assimilates into biomass

```{r wofost_crop_70, echo=FALSE}
p <- params[params[,1] == "conversion of assimilates into biomass", -1]
rownames(p) <- NULL
knitr::kable(p)
```


### maintenance respiration

```{r wofost_crop_80, echo=FALSE}
p <- params[params[,1] == "maintenance respiration", -1]
rownames(p) <- NULL
knitr::kable(p)
```

### partitioning

```{r wofost_crop_90, echo=FALSE}
p <- params[params[,1] == "partitioning", -1]
rownames(p) <- NULL
knitr::kable(p)
```

### death rates

```{r wofost_crop_100, echo=FALSE}
p <- params[params[,1] == "death rates", -1]
rownames(p) <- NULL
knitr::kable(p)
```

### water use

```{r wofost_crop_110, echo=FALSE}
p <- params[params[,1] == "water use", -1]
rownames(p) <- NULL
knitr::kable(p)
```

### rooting

```{r wofost_crop_120, echo=FALSE}
p <- params[params[,1] == "rooting", -1]
rownames(p) <- NULL
knitr::kable(p)
```

### CO2

```{r wofost_crop_130, echo=FALSE}
p <- params[params[,1] == "CO2", -1]
rownames(p) <- NULL
knitr::kable(p)
```

