# Example run

## Introduction

With WOFOST, crop growth is simulated using daily weather data parameters describing the crop(s) and soil type(s) of interest, as well as the crop management (such as planting date). You can calculate potential and water-limited production. In the current version, nutrient-limited production and yield reducing factors are not taken into account. In each model run, WOFOST simulates the growth of a specific crop, given the selected soil and weather data. For each simulation, you must select initial conditions such as the as the soil's water status, and management decisions such as the crop planting date.

## Chapter requirements

We use *R* package `Rwofost`. has been submitted to CRAN (March 2020), so pretty soon you should be able  [install](/installation.html) it from there.

## Input data
 
To run WOFOST, you need daily [weather data](weather.html). Here we use an file that comes with the `meteor` package.

```{r wofost_0}
library(Rwofost)
f <- system.file("extdata/Netherlands_Swifterbant.csv", package="meteor")
wth <- read.csv(f)
wth$date <- as.Date(w$date)
head(wth)
```

A variety is described by a (rather lengthy) set of [crop parameters](crop.html). The `Rwofost` package comes with a set of examples that you can use or build on. Let's take barley

```{r wofost_11}
crop <- wofost_crop("barley")
```

We can also use example [soil parameters](soil.html).

```{r wofost_20}
soil <- wofost_soil('ec1')
```

And "[control](control.html)" parameters that set things like planting date. 

```{r wofost_30}
contr <- wofost_control()
contr$modelstart <- as.Date("1976-02-06")
```

## Model

Now that we have the input data we can either run the model in one step 

```{r wofost_42}
out <- wofost(crop, wth, soil, contr)
```

Or in two steps, by first creating the [model](model.html).

```{r wofost_43}
m <- wofost_model(crop, wth, soil, contr)
```

And then running it 

```{r wofost_45}
d <- run(m)
```

A quick look at the [output](output.html).

```{r wofost_50}
head(d)
tail(d)
plot(d[,'step'], d[, 'LAI'])
```

