# Soil parameters

## Introduction 

## List of soils

Here is a list of the default soil parameter sets that come with Rwofost

```{r wofost_soil1}
library(Rwofost)
soils <- wofost_soil()
cat(soils, "\n")
```

You can get parameters for a single soil by like this. 

```{r wofost_soil10}
soil <- wofost_soil("ec1")
```

You can also specify a filename to read parameters from a file you created yourself. Have a look at the files in the Rwofost package. On your computer they are installed on what is returned by *R* if you run this: `system.file("wofost/soil", package="Rwofost")`

The object `soil` is a `list` with `r length(soil)` elements (some of the parameters are used to compute nutrient limitation but that has not been implemented yet --- so these can be ignored for now). 

```{r wofost_soil12}
class(soil)
length(soil)
str(soil[1:5])
```

Most are single numbers, but there are also two-row matrices (referred to as tables in the context of WOFOST). The names of these matrix-parameters end on TAB (for table). Here is an example. 

```{r wofost_soil14}
soil$SMTAB
```

## Parameters
	
### Soil water retention


```{r wofost_soil_20, echo=FALSE}
params <- read.csv("soilpars.csv", stringsAsFactors=FALSE)
params[,4] <- gsub("-1", "^-1^", params[,4])
p <- params[params[,1] == "soil water retention", -1]
rownames(p) <- NULL
knitr::kable(p)
```

### Hydraulic conductivity


```{r wofost_soil_40, echo=FALSE}
p <- params[params[,1] == "hydraulic conductivity", -1]
rownames(p) <- NULL
knitr::kable(p)
```

### Water infiltration and surface storage

```{r wofost_soil_50, echo=FALSE}
p <- params[params[,1] == "Water infiltration and surface storage", -1]
rownames(p) <- NULL
knitr::kable(p)
```

