```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Market access

*Sebastian Palmas*, [CIMMYT](https://www.cimmyt.org/)


## Editor's Notes 

This shows a lot of interesting spatial operations. These need more explanation througout.
You have create a travel cost surface. You can use that to compute market access. Perhaps with the gdistance package.

The decisions you have made need more discussion, explanation (e.g. units) and justification where possible. Please cite the literature. E.g. Andy Nelson's work. 

And perhaps follow that approach more closely?

Also see some comments in the text.

And I may have made changes that are not correct --- the final result does not look correct --- please fix were need be.

## Introduction

Many rural producers often face serious difficulties in accessing markets to sell their goods in the marketplace. They are constrained by their remote location and high transportation costs. We will estimate market access for Tanzania considering road access and land cover classes. 

## Chapter requirements.

For this chapter you need the following `R` packages: `terra`, `geodata`, and `reagro`. See [these instructions](/installation.html) about installing these packages.

We first create a cost surface for transportation. 

## Transportation cost surface 

### Elevation and Slope

The first factor we consider is slope. Slope is an important factor in travel costs. Flatter terrain is easier to travel through than sloping areas. We use the `geodata` package to download elevation data and then create a slope layer using the `terra::slope` function. 

```{r slope, message=FALSE}
library(geodata)
tza_alt <- elevation_30s("Tanzania", path=".")
tza_slope <- slope(tza_alt, unit="radians", neighbors=8)
plot(tza_slope, main="Slope")
```

We use the slope layer obtained above to create a decay coefficient that will be appied to each grid cell in the cost surface.

We use a decay coefficient of 1.5. This governs how much the slope impacts the speed

EXPLAIN MORE


```{r}
decay <- 1.5
slope_cost <- exp( decay * tan(tza_slope) )
names(slope_cost) <- "slope_cost"
```

We will the slope layer as template for the other rasters that we will create.


### Roads

#### Open Street Map data

Ed: Perhaps this section should be a recipe on its own.


Road access is important predictor of market access.

Market access estimates should account for road distance from a point. However, not all roads are equal. You can travel faster over a highway, and probably more cheaply per unit distance, than over a dirt road.

A good source of roads data is Open Street Map (OSM). (expand what it is, and limitations (completeness?))

You can download OSM data in *R* using the [osmdata](https://cran.r-project.org/web/packages/osmdata/index.html) package. You can only download a certain amount of data in a single request, so typically you need to a number of requests for adjacent regions and then combine the results. We used the first level administrative boundaries of Tanzania (also downloaded with the geodata package) for this. 

```{r gadm, message=FALSE, warning=FALSE}
adm <- gadm("TZA", level = 1, path=".")
```

We wanted data for primary, secondary and tertiary highways in Tanzania. Below is a helper function to accomplish this. 

```{r rdtp}
library(osmdata)

roadtypes <- c("primary", "secondary", "tertiary")

getOSMdata <- function(obj, key, value) {
    b <- sp::bbox(obj)
    q <- osmdata::opq(b)
    q <- osmdata::add_osm_feature(q, key, value)
    d <- osmdata::osmdata_sp(q)
    d$osm_lines[, key]
}

```

The below takes a long time to run. It may also fail because of the relative fraility of the OSM servers (or rather then enourmous number of requests that they get).

```{r getOSM, eval=FALSE}
x <- lapply(seq_along(adm), function(i) getOSMdata(adm[i, ], "highway", roadtypes))
```

Now put the results together 


```{r processOSM, eval=FALSE}
# combine the results for each area
rd <- do.call(bind, x)
# remove anything outside of Tanzania
rd <- crop(rd, adm)
# aggregate by type, also removing duplicates line parts
roads <- aggregate(rd, "highway")
```

The OSM data is very detailed. Much more than we need in most analysis. We simplified it with the `rmapshaper` package to store less data and so that the example runs faster. 

```{r simplr, eval=FALSE}
library(rmapshaper)
roads <- ms_simplify(roads, keep=0.01)
```
The results was put in the reagro package, and that is where we get it from

```{r plotRoads}
library(reagro)
roads <- reagro_data("tza_roads")
plot(tza_slope)
lines(roads)
lines(roads[roads$highway == "secondary", ], lwd=2, col="blue")
lines(roads[roads$highway == "primary", ], lwd=4, col="red")
```


#### Road cost surface

The road types have different moving speeds and, therefore, costs. Primary roads have faster speeds and lower travel costs than secondary and tertiary roads.

We rasterize the roads. The values of the rasters reflect the travel costs of moving through a cell by road (if there is one). We use the slope layer as template for the rasterization.

```{r rasterize}
cfile <- "rdcost.tif"

if (!file.exists(cfile)) {

	i <- match(roads$highway, roadtypes)
	roads$speed <- c(0.001, 0.0015, 0.002)[i]
	
	rd_cost <- rasterize(roads, tza_slope, values="speed", filename=cfile, wopt=list(names="slope_cost"))
	
} else {
	# no need to repeat if we already have done this
	rd_cost <- rast(cfile)
} 

plot(rd_cost)
```


### Land Cover

Land cover affects the access of a certain point. A point inside a closed forest has lower accessibility than a point within croplands or within bare areas.

In this example, we will use the GLOBCOVER 2009 Version 2.3 classification for Tanzania. 
INSERT URL

We included these data the reagro package.

To In order to be able to analyze all layers, we need to have them all in the same projection and resolution. We will use the slope layer as the template for all layers.

```{r}
tza_lc <- reagro_data("TZA_globcover")
plot(tza_lc, main = "GLOBCOVER 2009 v2.3 land classes")
lines(adm)
```

Creating a travel cost surface depending on land cover requires an associated travel cost for each land class.

Below we create a table to assign travel cost values to each land cover class. 

```{r}
rc <- data.frame(from=as.vector(unique(tza_lc)), to=0.02)
rc$from[rc$from %in% c(190,200)] <- 0.01
rc$from[rc$from == 160] <- 0.03
rc$from[rc$from %in% c(40,50,70)] <- 0.04
rc$from[rc$from == 170] <- 0.05
rc$from[rc$from == 210] <- 0.11
rc$from[rc$from == 220] <- 0.13

#reclassifying
tza_lc_cost <- classify(tza_lc, rc) 
```


```{r}
lcfname <- "lc_cost.tif"
if (!file.exists(lcfname)) {
  lc_cost <- warp(tza_lc_cost, tza_slope, filename=lcfname, wopt=list(names="lc_cost"))
} else {
  lc_cost <- rast(lcfname)
}
```


```{r}
plot(lc_cost, main = "Travel costs depending on land cover class")
```


### Combining all travel costs

Now that we have all travel cost surfaces what we will use, we then need to combine them into a single travel cost layer that keeps only the minimum cost associated of moving though each grid cell. In this case, we combine the three cost layers in a SpatRaster object and apply the `min` function to obtain the minimum value in each grid cell.


```{r}
# Combine the cost layers
all_cost <- c(rd_cost, lc_cost, slope_cost)

#getting the minimum value of each grid cell
cost <- min(all_cost, na.rm=TRUE)

plot(cost, main="Final cost layer")
```


## Market access

With the cost surface, we can compute market access .... 


## References



## Citation 

Palmas, S., 2019. Spatial model of market access. In: Hijmans, R.J. and J. Chamberlin. Regional Agronomy: a pratical handbook. CIMMYT. https:/reagro.org/recipes/access.html

