# More Milk in Tanzania

The MoreMilkiT project, funded by ILRI in collaboration with multiple stakeholders and partners, is working in Tanzania to secure income for marginalized dairy producing communities through dairy market hubs. The project has three objectives: to develop value chains, generate evidence to increase participation in dairy value chains, and inform policy. MoreMilkiT has conducted two surveys to date: the baseline survey, which interviewed 932 households, and the monitoring survey, whih interviewed 461 households. Following the baseline survey, ILRI published a research brief with preliminary results to assess the progress to date in establishing dairy market hubs. This case study will assess and summarize the baseline data using the research brief. The data used in this case study can be downloaded from ILRI [here](http://data.ilri.org/portal/dataset/moremilkit-hh-tz).

## Baseline data 

```{r data, warning = F}
datapath <- "../data/MoreMilkiT/baseline/baseline_stata"
ff2 <- list.files(datapath, pattern='\\.dta$', full=TRUE)
library(readstata13)
x2 <- lapply(ff2, read.dta13)

z2 <- strsplit(basename(ff2), '_|[.]')
z2 <- t(sapply(z2, function(x) x[1:8]))
z2[z2=='dta'] <- ""
z2[is.na(z2)] <- ""
z2 <- apply(z2, 1, function(i) paste(i, collapse=""))
names(x2) <- z2

```

Luckily the dataset includes the questionnaire used during interviews, which we can refer to for help in finding individual variables and locations of tables. We will begin with basic summary statistics. In this section we find the number of cattle keeping households in each district. 

```{r Bydistrict}
#extract information on household id and whether the household has cattle
cattle <- x2$SEC1HOUSEHOLDIDENTIFICATION[,c("tanzania_dp_id", "cattle_keepers")]
cattle$tanzania_dp_id <- as.character(cattle$tanzania_dp_id)
cattle$tanzania_dp_id <- substr(cattle$tanzania_dp_id, 0, 1)

#Number of cattle keeping households per district (from table 2)
bydistrict <- aggregate(cattle$cattle_keepers, list(cattle$tanzania_dp_id), sum, na.rm=T)
bydistrict$Group.1 <- c("Lushoto", "Mvomero", "Handeni", "Kilosa")
colnames(bydistrict) <- c("District", "No. of cattle keeping households")
knitr::kable(bydistrict)
```

We can create a dataframe with basic information about the household heads from all ofthe districts. We chose certain variables to include in this dataframe based on summary information shown in the report, including age and education information. 

```{r age and education of all}
all <- x2$SEC2HOUSEHOLDROSTER[x2$SEC2HOUSEHOLDROSTER$relationship_to_hh_head=="Head",]
all$age_years[all$age_years<0] <- NA
#To convert the id into just district information
all$tanzania_dp_id <- as.character(all$tanzania_dp_id)
all$tanzania_dp_id <- substr(all$tanzania_dp_id, 0, 1)
all$tanzania_dp_id <- as.numeric(all$tanzania_dp_id)
all$illit <- ifelse(all$education_level=="No formal and illiterate", 1, 0)
all$noformal <- ifelse(all$education_level=="No formal but literate", 1, 0)
all$primary <- ifelse(all$education_level=="Primary school", 1, 0)
all$post <- ifelse(all$education_level=="High/secondary" | all$education_level=="College" | all$education_level=="University", 1, 0)
```

Next, we represent table 3 and 4 of the baseline report. These tables show the age and education attainment of male and female household heads. 
```{r Age and edu of male}
male <- all[all$gender==1,c(1, 6, 16:19)]
#household statistics by district
maledist <- aggregate(male[2:6], list(male$tanzania_dp_id), mean, na.rm=T)
#total statistics for male houeshold
maletot <- as.data.frame(apply(male, 2, mean, na.rm=T))
maledist[5,] <- t(maletot)
maledist[,3:6] <- maledist[,3:6]*100
maledist <- round(maledist, 1)
maledist$Group.1 <- c("Lushoto", "Mvomero", "Handeni", "Kilosa", "Total")
colnames(maledist) <- c("District", "Age", "No formal and illiterate", "No formal but literate", "Primary school", "Post Primary")
kable(maledist, caption = "Age and Education of Male Household Heads")
```

```{r age and edu of female}
female <- all[all$gender==2,c(1, 6, 16:19)]
#stats by district
femaledist <- aggregate(female[2:6], list(female$tanzania_dp_id), mean, na.rm=T)
#total stats female
femaletot <- apply(female, 2, mean, na.rm=T)
#combine to single table
femaledist[5,] <- t(femaletot)
#convert edu to percentages
femaledist[,3:6] <- femaledist[,3:6]*100
#Round variables
femaledist <- round(femaledist, 1)
#Rename the first column
femaledist$Group.1 <- c("Lushoto", "Mvomero", "Handeni", "Kilosa", "Total")
#rename the colnames
colnames(femaledist) <- c("District", "Age", "No formal and illiterate", "No formal but literate", "Primary school", "Post Primary")
kable(femaledist, caption= "Age and Education of Female Household Heads")

```

Another summary statistic reported is the pattern in milk sales across months. In section 6 of the survey, participants are asked whether or not they sell milk in each of the 12 months. We can count the number of househlds that repond with 'yes', using the function sapply below, and then create a dataframe of the count of sales in each month.  
```{r months sold fresh milk}
milksold <-x2$SEC64FRESHMILKSALE
milksold$tanzania_dp_id <- as.character(milksold$tanzania_dp_id)
milksold$tanzania_dp_id <- substr(milksold$tanzania_dp_id, 0, 1)

Mvomero <- milksold[milksold$tanzania_dp_id=="2",]
Handeni <- milksold[milksold$tanzania_dp_id=="3",]
Kilosa <- milksold[milksold$tanzania_dp_id=="4",]
lus <- milksold[milksold$tanzania_dp_id=="1",]

milklus <- sapply(lus[,5:16],FUN = function(x){length(x[x=="Yes"])})
milkmvo <- sapply(Mvomero[,5:16],FUN = function(x){length(x[x=="Yes"])})
milkhan <- sapply(Handeni[,5:16],FUN = function(x){length(x[x=="Yes"])})
milkkil <- sapply(Kilosa[,5:16],FUN = function(x){length(x[x=="Yes"])})
milkall <- data.frame(milklus, milkmvo, milkhan, milkkil)
```

Once we have the dataframe, we can plot the milk sales as is done in figure 2 of the baseline report. 
source: https://www.harding.edu/fmccown/r/ 

```{r Plot}
#Initial plot of milk sales in Lushoto
plot(milkall$milklus, type="o", pch=18, col="blue", ylim=c(20,100), axes=F, ann=F)
#set up x axis
axis(1, at=1:12, lab=c("Nov", "Dec", "Jan", "Feb", "Mar", "April", "May", "June", "July", "Aug", "Sep", "Oct"))
#set up y axis
axis(2, las=1, c(20, 40, 60, 80, 100))
#put box around graph
box()
#add additional lines for the additional 3 districts
lines(milkall$milkmvo, type="o", pch=15, col="red")
lines(milkall$milkhan, type="o", pch=17, col="green")
lines(milkall$milkkil, type="o", pch=4, col="purple")
#add titles
title(main="Annual Pattern in Milk Sale")
title(xlab="Month")
title(ylab="Count of dairy households selling milk")
#add legend
legend("bottom",c("Lushoto", "Mvomero", "Handeni", "Kilosa"), col=c("blue", "red", "green", "purple"), pch=c(18, 15, 17,4), horiz=T)
```

