
# MODIS data pre-processing

In this chapter we download, pre-process and compute NDVI from MODIS data. 
We discussed these procedures in detail [here](https://gfc.ucdavis.edu/events/agrin/html/modis/index.html). 

## Area of interest

First we get the area of interest. Normally you would get this from a file (e.g. a shapefile). See `?terra::vect` for reading such files.

```{r aoi, message=FALSE}
library(reagrodata)
aoi <- data_ibli("marsabit") 
aoi
```

We can see that we have 58 polygons, representing 58 unique "sub-locations". Let's plot the polygons on top of a map of Kenay what we have.

```{r ken, message=FALSE}
library(geodata)
ken <- gadm("KEN", level=1, path=".")
plot(ken, border="gray")
plot(ken[ken$NAME_1 == "Marsabit", ], col="beige", border="gray", add=TRUE)
lines(aoi, col="red")
```

Clearly in northern Kenya. Note that some polygons are much smaller than others.



## Get MODIS files

Here we show how to download and pre-process the data, but only for 1 month, to limit the download and processing needed. After that we use pre-processed data.

Set up the download directory

```{r downloaddir}
datadir <- file.path(dirname(tempdir()), "agrins/ibli/modis")
dir.create(datadir, recursive=TRUE, showWarnings=FALSE)
# datadir
```

In this example we download data from 2010 only. 

```{r dates}
start <- "2010-01-01" 
end <- "2010-01-07"
```


Now download the data

```{r moddown, message=FALSE}
library(luna)
fmod <- getModis(product="MOD09A1", start, end, aoi, download=TRUE, path=datadir)
fmod
```

We only have two files here, each representing 8 days ---- just to illusrate the processing.

A single MODIS tile for extends well over the study area. We first crop the data to correspond to the study area, and then we remove bad pixels (clouds and shadow) and then .

## Quality

To remove bad quality pixels from the selected files, we follow the same method described [here](https://rspatial.org/terra/modis/index.html). First, specify matrix "se" with start and end of the quality assessment (QA) bits considered, and specify a list "reject" with the values to be rejected (in the image) matching the rows in "se".

```{r qcMat}
se <- matrix(c(1,2,3,6,11,11,12,14,16,16), ncol=2, byrow=TRUE)
reject <- list(c("10", "11"), c("1100","1101","1110","1111"), "1", c("000","110","111"), "11")
```

Project case study boundary to MODIS sinusoidal projection system.

```{r prj}
prj <- "+proj=sinu +lon_0=0 +x_0=0 +y_0=0 +a=6371007.181 +b=6371007.181 +units=m +no_defs "
pols <- project(aoi, prj)
```

## Monthly NDVI computation 

We now compute NDVI for each month. First, produce a quality mask, mask out bad quality pixels, crop to the area of interest, and then compute NDVI. 

We use a loop over all files (`fmod`).

```{r Preprocess}
for(i in 1:length(fmod)) {
  # Load a MODIS file
  r <- rast(fmod[i])

  # Generate quality mask (layer 12)
  quality_mask <- modis_mask(r[[12]], 16, se, reject)

  #Select only red and NIR
  r <- r[[1:2]]
  names(r) <- c("red", "NIR")

  # remove low quality pixels
  r <- mask(r, quality_mask)

  # remove areas outside the AOI rectangle
  r <- crop(r, pols)

  # ensure that all values are between 0 and 1 
  r <- clamp(r, 0, 1) 

  # Compute NDVI
  ndvi <- (r$NIR - r$red)/(r$NIR + r$red)

  filename <- gsub(".hdf$", "_ndvi.tif", fmod[i])
  writeRaster(ndvi, filename=filename, overwrite=TRUE)
}
```

