# Spatio-temporal aggregation

## Temporal NDVI aggregate
 
For the entire analysis we use precomputed NDVI for the period between 2008 to 2015. The next code block downloads the pre-processed data. The first time this takes a while. After that, it should be quick as the data won't be downloaded again if they are still on your disk.

```{r NDVIdownload, message=FALSE}
datadir <- file.path(dirname(tempdir()), "ibli/modis")
dir.create(datadir, recursive=TRUE, showWarnings=FALSE)
files <- agrodata::data_ibli("marsabit_modis_ndvi", datadir)
head(basename(files))
```

Get the area of interest again. We project to the sinusoidial coordinate refrence system that the MODIS data are in 

```{r aoi}
aoi <- data_ibli("marsabit")
sinprj <- "+proj=sinu +lon_0=0 +x_0=0 +y_0=0 +a=6371007.181 +b=6371007.181 +units=m +no_defs "
pols <- project(aoi, sinprj)
```

Now set up some parameters to select the subsets of files that we need.

```{r NDVIs}
# Define time period
startyear <- 2008
endyear   <- 2015

# Define a season (LRLD in this case)
sSeason <- "LRLD"
startSeason <- "03-01"
endSeason   <- "09-30"
```

Below we define a function for selecting MODIS files for a given year and season.

```{r selectModFs, message=FALSE}
library(luna)
selectModisFiles <- function(files, startdate, enddate) {
  base_names <- basename(files)
  dates <- substr(base_names, 10, 16)
  dates <- dateFromYearDoy(dates)
  i <- (dates >= as.Date(startdate)) & (dates <= as.Date(enddate) )
  files[i]
}
```

Let's try the function. First for a single year, and to make a plot

```{r NDVIs_Plots, fig.width=10, fig.height=10} 
  season <- selectModisFiles(files, paste0(startyear, "-", startSeason), paste0(startyear, "-", endSeason))
  sndvi <- rast(season)
  plot(sndvi[[1:9]])
```


We can compute 8-day NDVI spatial aggregates by region.

```{r NDVIs_Plots2} 
ndvi_val <- terra::extract(sndvi, pols, fun=mean, na.rm=TRUE)
# the first region
plot(ndvi_val[1,], ylab= "NDVI")
```


Now, we process all files

```{r UseselectModFs}
workdir <- file.path(dirname(tempdir()), "ibli/work")
dir.create(workdir, recursive=TRUE, showWarnings=FALSE)

for(year in startyear:endyear) {
  season <- selectModisFiles(files, paste0(year, "-", startSeason), paste0(year, "-", endSeason))
  
  sndvi <- rast(season)
  ndvimean <- mean(sndvi, na.rm = TRUE)
  
  # Shorten layer names 
  names(sndvi) <- paste0("ndvi_",dateFromYearDoy(substr(names(sndvi), 10, 16)))
  
  filename=file.path(workdir, paste0( 'MOD09A1.h21v08.',year,"_",sSeason,'_ndvit.tif'))

  writeRaster(ndvimean, filename = filename , overwrite=TRUE)
}
```


Here is an example of the files we have created.

```{r MorePlots}
par(mar = c(2.2, 2.2, 1.2, 0.5)) #c(bottom, left, top, right)
stitle <- paste(year, " NDVI", sep="")
plot(ndvimean, main = stitle)
```


## Spatial aggregate

Above we computed the seasonal mean NDVI for each pixel. Next we compute spatially aggregated NDVI for each sub-location.


```{r NDVISpatial}
season <- "LRLD"
files <- list.files(workdir, pattern=paste0(season,"_ndvit.tif$"), full.names=TRUE)

output <- matrix(nrow=nrow(pols), ncol=length(files))
colnames(output) <- substr(basename(files),16,19)

for(i in 1:length(files)){
	# read the NDVI rasters for each season
	ndvi <- rast(files[i])
	# Compute spatial aggregates by sub-location
	output[,i] <- extract(ndvi, pols, fun=mean, na.rm=TRUE)
}
```


Now we combine the aggregated data with the polygon data, and we save it to disk as we might want to use it again later.


```{r comb}
# Create a data.frame
res <- data.frame(SLNAME=pols$SUB_LOCATION, IBLI_Zone=pols$IBLI_UNIT, output, stringsAsFactors = FALSE, check.names=FALSE)

saveRDS(res, file.path(workdir, paste0(season,"_ndvi_st_mat.rds")))
```

Now we can plot the seasonal mean NDVI over time, by region.

```{r NDVIls_Plots}
plot(output[1,], ylab="NDVI", xlab="Year", type="l", ylim=c(0,.6))
for (i in 2:nrow(output)) {
  lines(output[i,])
}

```


## Z-scores

Now we can compute z-scores based on the period 2008--2015 as an example. Z-scores indicate the deviations of seasonal *NDVI* from its longterm annual mean. The unit is "standard deviations". This a z-score of -1 means that the value is 1 standard deviation below the (expected) mean value.

z-score computation function:

```{r zscore}
zscore <- function(y){
  (y - mean(y, na.rm=TRUE) ) / (sd(y, na.rm=TRUE))
}

```

Now we can use this function to compute z-scores over the LRLD period. You can also make some plots to assess forage availablity. For instance, assume a threshold of zero and make payouts anytime the zonal mean deviates below the historical average. In addition, instead of simple line plot you can use linear regression and some local smoothing functions like moving average. 

```{r zScores}
scores <- apply(res[,-c(1:2)], 1, zscore)
scores <- t(scores)
scores[1:5, 1:6]

ndvi_st <- data.frame(SLNAME=pols$SUB_LOCATION, IBLI_Zone=pols$IBLI_UNIT, scores, stringsAsFactors = FALSE, check.names=FALSE)
saveRDS(ndvi_st, file.path(workdir, paste0(sSeason,"_zndvi_st_mat.rds")))

```


We can show z-scores over the analysis years per sub-region. Here for four regions

```{r PlotzScores, fig.width=12, fig.height=12}
years <- 2008:2015

par(mfrow=c(2,2))
for(i in 1:4){
  
	plot(years, scores[i,], ylab="zNDVI", xlab="Year", main=pols$SUB_LOCATION[i])
	lines(years, scores[i,])

# Line showing where NDVi is 1 sd below the mean
	abline(h=-1, col="red",lty=2)
}
```


