# Evaluation


## Introduction

Use the insurance scheme designed in the previous [chapter](6-design.html). 

First load the mortality predictions data.

```{r eval10}
d <- readRDS("pred_mort1.rds")
d <- na.omit(d)
```

Compute mortality losses as the difference between the observed mortality (*mortality_rate*) and the trigger we adopted previously. Subsequently, compute insurance pay as estimated from predicted mortality (*predicted_mortality*) using the same trigger.

```{r eval11}
trigger <- 0.20
d$loss <- pmax(0, d$mortality_rate - trigger)
d$pay  <- pmax(0, d$predicted_mortality - trigger)
head(d, n=2)
```

Below is a plot of pay as computed from predicted mortality versus observed. Ideally, if predictions are the same as observed then their plot should form a straight line.

```{r eval12}
plot(d$loss, d$pay, xlab="Loss", ylab="Pay")
abline(0,1, col="red")
```

What do you make of the plots? What of a plot of payouts versus *zNDVI*?

```{r ev1}
plot(pay~NDVI, data=d, col="red")
```

We now need to assess the welfare of index insurance to the pastoralists. We use the certainty equivalent function.

```{r util}
library(agro)
agro::ce_income
```

Let us compute the premium that is paid for IBLI insurance as equivalent to the expected payouts with a mark up of 25%. We will basically compare the IBLI welfare based on the observed mortality (*base*) and predicted mortality (*insurance*).

```{r eval20}
markup <- 0.25
premium <- mean(d$pay) * (1 + markup)
premium

base <- 1-d$mortality_rate
insurance_nomarkup <- base + d$pay
insurance <- insurance_nomarkup - premium

base <- base * 100
insurance <- insurance * 100
insurance_nomarkup <- insurance_nomarkup * 100
```

Compute CE with insurance (*insurance*) and without (*base*) respectively and compare them based on $\rho$.

```{r eval21}
rho <- 2
ce_base <- ce_income(base, rho)
ce_ins25  <- ce_income(insurance, rho)
ce_ins_nomarkup  <- ce_income(na.omit(insurance_nomarkup), rho)

ce_base 
ce_ins25
ce_ins_nomarkup

ce_ins25 / ce_base
ce_ins_nomarkup/ce_base
head(d, n=2)

```

Compute CE over $0 \leq \rho \leq 10$ and plot CE pay percentage when there is no mark up on insurance (*ce_ins_nomarkup*), when insurance is 25% marked up (*ce_ins*) and a case where no insurance is taken by pastrolist (*ce_base*).

```{r eval22}
rhos <- seq(0, 3, .1)
ce_base <- ce_ins <- ce_ins_nomarkup <- rep(NA, length(rhos))
for(i in 1:length(rhos)){
  ce_base[i] <- ce_income(base, rhos[i])
  ce_ins[i]  <- ce_income(insurance, rhos[i])
  ce_ins_nomarkup[i] <- ce_income(insurance_nomarkup, rhos[i])
}

inc <- seq(0.1, 1, 0.1)
plot(rhos, ce_base, type="l", col="blue", ylab= "Pay (%)", xlab="CRRA", cex=2, ylim=c(70, 90))

lines(rhos, ce_ins, col="red")
lines(rhos, ce_ins_nomarkup, col="green")

legend("topright", c("No insurance", "25% Marked-up Insurance", "No mark up"), lty=1, col=c("blue", "red", "green"), title = "IBLI contract type", bty = "n")

```

What do you observe from the graphs? What happens when you change the trigger to say 20%?

Let do the MQS test using the same values of CRRA and make a plot to determine pastrolist welfare with and without insurance.

```{r eval23}
ce_base <- ce_ins <- mqs <- rep(NA, length(rhos))

for(i in 1:length(rhos)){
  ce_base[i] <- ce_income(base, rhos[i])
  ce_ins[i]  <- ce_income(insurance, rhos[i])
  mqs[i] <- ce_ins[i] - ce_base[i]
}

inc <- seq(0.1, 1, 0.1)
plot(rhos, mqs*100, type="l", col="blue", ylab= "MQS (insurance benefit in %)", xlab="CRRA", cex=2)
abline(h=0, col="red")

```

From the plot you can observe that pastoralists with a risk aversion less than 1.5 do not derive any value from the insurance. However, those with higher risk aversion derive more benefit from the insurance as the fall above the zero line.

## Simulation

Remember we predicted mortality previously based on *zNDVI* however there is uncertainty that is associated with our model. We will therefore conduct a simulation of pay based on 95% confidence interval of the model's predictions.

Let us set up the environment including required library and data.

```{r eval100, message=FALSE}
library(msir)
dd <- na.omit(d)

```

Create a model that predicts mortality from *zNDVI*. 

```{r eval200}
m <- loess.sd(dd$NDVI, dd$mortality_rate)
fitsd <- cbind(fit=m$model$fitted, sd=m$sd)

```

Let us sample *1000* times from the model we created previously.

```{r eval3}
ns <- 1000
sample <- apply(fitsd, 1, function(i) {
  pmin(1, pmax(0, rnorm(ns, i[1], i[2])))
})
sample <- t(sample)
```

Compute insurance premium with 25% and with no markup from the samples

```{r eval4}
markup <- 0.2
premium <- mean(dd$pay) * (1 + markup)
```

and their corresponding CE at $\rho=2$.

```{r eval6}
out_nomarkup <- out_ins <- out_base <- rep(NA, ns)
rho = 2

for (i in 1:ns) {
  base <- 1- sample[,i]
  insurance_nomarkup <- base + dd$pay
  insurance <- insurance_nomarkup - premium
  out_base[i] <- ce_income(100*base, rho)
  out_ins[i]  <- ce_income(100*insurance, rho)
  out_nomarkup[i]  <- ce_income(100*insurance_nomarkup, rho)
}

mean(out_base)
mean(out_ins)
mean(out_nomarkup)

```

Make some plots.

```{r eval5}
benefit_ins <- out_ins - out_base
benefit_nomu <- out_nomarkup - out_base
b <- cbind(benefit_ins, benefit_nomu)
boxplot(b, ylim=c(-1,10))
```

The graph illustrates that the contract passes the welfare MQS test. 

